---
title: "Assignment 2 Analysis: BMI by education among NHANES adults"
author: Morgan Ehman
format: html
---

## Setup

```{r}
#| message: false
#| warning: false
library(tidyverse)  # AI-EDIT(2026-06-23): tidyverse-default — consolidated core library() calls into library(tidyverse)
```

## Load Data

```{r}
nhanes <- read_csv("examples/nhanes-equity/data/nhanes_equity_v6.csv", show_col_types = FALSE)
```

## Inspect

```{r}
dim(nhanes)
names(nhanes)
glimpse(nhanes)
colSums(is.na(nhanes))
```

The data have 105,626 rows and 18 columns. Education is missing in 47,251 records, mostly among participants younger than 20, 
while BMI is missing in 9,356 records.
AI NOTE: I used Claude (Opus 5.5) to suggest possible summary analyses. 
It suggested BMI by Education for adults as one option, given Education is only recorded for adults, i.e., 
individuals aged 20 years or older. On my visual skim, this appeared to be true, but I wanted to verify. 

## Verify education missingness and non-adulthood (i.e., individuals with age < 20)

```{r}
# AI prompt: "Give me a line or two of code to check that Education is only recorded for adults."
# Verified: youngest age with education recorded is 20; all 47,164 non-adults are missing education.

# Check youngest age with education recorded (expect it to be 20)
min(nhanes$Age[!is.na(nhanes$Education)])

# Check Education missingness between adults and non-adults (expect it to be missing for all non-adults)
nhanes |>
  group_by(adult = Age >= 20) |>
  summarise(n = n(), missing_education = sum(is.na(Education)))
```

The youngest individual with Education recorded is 20 years old, and all 47,164 non-adults (i.e., individuals with Age < 20) are missing Education.

## Summarize

AI NOTE: Once I verified that Education is only recorded for adults, I prompted Claude to 
summarize BMI by Education for adults (individuals 20+), note individuals with a BMI >=30, and 
to show missing values instead of dropping them. 

```{r}
# AI prompt: "Summarize BMI by Education for adults 20+, add a BMI 30+ flag with mutate(),
# and show missing BMI in each group instead of dropping it."
# Verified: College Grad row recalculated separately and matched; NA row matches the
# 87 adults missing Education; group sizes add up to all adults (see 'Verify Summary' section below).

adults <- nhanes |>
  filter(Age >= 20)

education_summary <- adults |>
  mutate(obese = BMI >= 30) |>
  group_by(Education) |>
  summarise(
    n             = n(),
    n_missing_bmi = sum(is.na(BMI)),
    mean_bmi      = round(mean(BMI, na.rm = TRUE), 1),
    pct_obese     = round(100 * mean(obese, na.rm = TRUE), 1),
    .groups       = "drop"
  )

education_summary
```

Among adults (i.e., individuals who are 20+) with recorded education, college graduates had the lowest mean BMI (27.9) 
and the lowest share with BMI 30 or higher (29.3%). Adults with some college had the highest mean BMI and proportion of 
individuals with BMI 30 or higher (29.6 and 40.5%, respectively), slightly above those with high school or less. The NA row 
holds the 87 adults with no Education recorded. These are unweighted, descriptive summaries pooled across survey cycles; they do 
not show that education causes differences in BMI.

## Verify Summary 

AI NOTE: I asked Claude for code to check that the summary table values are correct. The checks recalculate one row by a different route, 
confirm the NA row, and confirm no adults were lost.

```{r}
# AI prompt: "Give me code to check that the values in the education summary table are correct."
# Verified: every expected value below matched the table.

# Columns used in the summary exist (expect TRUE)
all(c("Age", "BMI", "Education") %in% names(nhanes))

# Recalculate the College Grad row: subset first, then calculate (compare with row 1)
college <- adults |> filter(Education == "College Grad")
 nrow(college)                                          # n (expect 13319)
 sum(is.na(college$BMI))                                # n_missing_bmi (expect 183)
 round(mean(college$BMI, na.rm = TRUE), 1)              # mean_bmi (expect 27.9)
 round(100 * mean(college$BMI >= 30, na.rm = TRUE), 1)  # pct_obese (expect 29.3)

# The NA row is the adults with no education recorded (expect 87)
sum(is.na(adults$Education))

# Check: no adults lost from the summary (expect TRUE)
sum(education_summary$n) == nrow(adults)
```

The summary table values were verified by recalculating the College Grad row, confirming the NA row, and confirming that no adults were lost from the summary. All checks passed.

## Function

AI NOTE: I asked Claude to suggest a simple function, and provided the original function as an example. 
It suggested a function that returns n, number missing, mean and SD for a numeric variable. 

```{r}
# AI prompt: "Suggest a simple function that returns n, number missing, mean, and SD
# for a numeric variable" (with the starter's mean_with_n() as an example).
# Verified: results for adults$BMI matched single-line calculations; known-answer test
# on made-up values passed (see 'Verify Function' section below).

describe_numeric <- function(x) {
  tibble(
    n         = length(x),
    n_missing = sum(is.na(x)),
    mean      = round(mean(x, na.rm = TRUE), 2),
    sd        = round(sd(x, na.rm = TRUE), 2)
  )
}

describe_numeric(adults$BMI)
describe_numeric(adults$Age)
```

The describe_numeric() function summarizes all 58,462 adults aged 20 or older. BMI is missing for 1,162 adults. Among adults with observed BMI, the mean is 29.0 and the standard deviation is 6.87. Age has no missing values, a mean of 50.1, and a standard deviation of 18.2.


## Verify Function

AI NOTE: To verify that the function was working as intended, I asked Claude to provide single lines of code 
that individually return n, number missing, mean, and SD for BMI, plus a known-answer test.

```{r}
# AI prompt: "Give me single lines of code that return n, number missing, mean, and SD
# for adults$BMI, to compare with describe_numeric()."
# Verified: all four values and the known-answer test matched.

# Each line should match describe_numeric(adults$BMI)
nrow(adults)                    # n (expect 58462)
sum(is.na(adults$BMI))          # n_missing (expect 1162)
mean(adults$BMI, na.rm = TRUE)  # mean (expect 29.037, which rounds to 29.0)
sd(adults$BMI, na.rm = TRUE)    # SD (expect 6.865727, which rounds to 6.87)

# Known-answer test on made-up values (expect n = 4, n_missing = 1, mean = 30, SD = 5)
describe_numeric(c(25, 30, NA, 35))
```

The describe_numeric() function was verified by comparing its output for adults$BMI with single-line calculations of n, number missing, mean, and SD. All values matched. A known-answer test on made-up values also passed.

## Save Output

```{r}
dir.create("assignments/assignment02-r-with-ai/submission/output", recursive = TRUE, showWarnings = FALSE)
write_csv(education_summary, "assignments/assignment02-r-with-ai/submission/output/summary_table.csv")
```

## AI Audit Note

I used Claude (Opus 5.5) to map the assignment requirements, suggest a summary question (BMI by Education among adults), 
and draft the summary table and the describe_numeric() function. I also used Claude to suggest code to verify the summary table and function. 
I verified all AI suggestions by checking the output and the checks for each. Each AI-assisted chunk has 
`# AI prompt:` and `# Verified:` comments. Further details are in `ai-use-note.md`.

